The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, and Substructure Searching

### 2016-08-15 article.tex
Two more sections to be added, based on content from John
Added Kai Dührkop as coauthor
Added more details to the molecular formula section, added a table with formula generator performance evaluation

% Load packages
\usepackage{url}  % Formatting web addresses  
\urlstyle{rm}
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amssymb}
%\usepackage{cite}
\usepackage{graphicx}

@@

\author[
   addressref={um},                   % id's of addresses, e.g. {aff1,aff2}
   %corref={aff1},                       % id of corresponding address, if any
   %noteref={n1},                        % id's of article notes, if any
   email={egon.willighagen@maastrichtuniversity.nl}   % email address
]{\inits{ELW}\fnm{Egon L} \snm{Willighagen}}
\author[
   addressref={aff2},
   email={john@nextmovesoftware.com}
]{\inits{JWM}\fnm{John W} \snm{May}}
\author[
   addressref={uppsala},
   email={jonathan.alvarsson@farmbio.uu.se}
]{\inits{JA}\fnm{Jonathan} \snm{Alvarsson}}
\author[
   addressref={uppsala},
   email={berg.arvid@gmail.com}
]{\inits{AB}\fnm{Arvid} \snm{Berg}}
\author[
   addressref={idea},
   email={jeliazkova.nina@gmail.com}
]{\inits{NJ}\fnm{Nina} \snm{Jeliazkova}}
\author[
   addressref={wi_mit},
   email={pluskal@wi.mit.edu}
]{\inits{TP}\fnm{Tomáš}~\snm{Pluskal}}
\author[
   addressref={???},
   email={miguelrojasch@googlemail.com}
]{\inits{MRJ}\fnm{Miguel}~\snm{Rojas-Cherto}}
\author[
   addressref={uppsala},
   email={ola.spjuth@farmbio.uu.se}
]{\inits{OS}\fnm{Ola} \snm{Spjuth}}
\author[
   addressref={???},
   email={gilleain.torrance@gmail.com}
]{\inits{GT}\fnm{Gilleain} \snm{Torrance}}
\author[
   addressref={um},
   email={chris.evelo@maastrichtuniversity.nl}
]{\inits{CTE}\fnm{Chris T} \snm{Evelo}}
\author[
   addressref={nih},
   email={guhar@mail.nih.gov}
]{\inits{RG}\fnm{Rajarshi} \snm{Guha}}
\author[
   addressref={ebi},
   email={steinbeck@ebi.ac.uk}
]{\inits{CS}\fnm{Christoph}~\snm{Steinbeck}}

\address[id=um]{
  \orgname{Dept of Bioinformatics - BiGCaT, NUTRIM, Maastricht University}, % university, etc
  %\street{Waterloo Road},                     %
  \postcode{NL-6200 MD},
  \city{Maastricht},                              % city
  \cny{The Netherlands}                                    % country
}
\address[id=aff2]{
  \orgname{??},
  %\street{D\"{u}sternbrooker Weg 20},
  %\postcode{24105},
  %\city{Kiel},
  \cny{UK}
}
\address[id=uppsala]{
  \orgname{Department of Pharmaceutical Biosciences, Uppsala University},
  %\street{D\"{u}sternbrooker Weg 20},
  \postcode{751 24},
  \city{Uppsala},
  \cny{Sweden}
}
\address[id=idea]{
  \orgname{Ideaconsult Ltd},
  \street{A. Kanchev 4},
  \postcode{1000},
  \city{Sofia},
  \cny{Bulgaria}
}
\address[id=nih]{
  \orgname{National Center for Advancing Translational Science},
  \street{9800 Medical Center Drive},
  \postcode{MD 20878},
  \city{Rockville},
  \cny{USA}
}
\address[id=ebi]{
  \orgname{Chemoinformatics and Metabolism team, European Bioinformatics Institute},
  %\street{A. Kanchev 4},
  %\postcode{1000},
  \city{Hinxton},
  \cny{UK}
}
\address[id=wi_mit]{
  \orgname{Whitehead Institute for Biomedical Research}, 
  \street{9 Cambridge Center},
  \postcode{MA 02142},
  \city{Cambridge},
  \cny{USA}
}
%\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
%\end{artnotes}

@@

The chemical formula is the basic/simple chemical representation of a compound.
It defines the number of isotopes or elements that constitute the composition
of a compound without describing how atoms are bonded. With the rise of
metabolomics it has become increasingly relevant to have full support for these
in cheminformatics libraries~\cite{RojasCherto2011}.

The CDK interfaces can handle several concepts related to chemical formulas:
the formula itself, sets of formulas, chemical formula ranges, adducts,
isotope containers and patterns, and rules to filter formula sets. These new
tools can be used for a number of tasks, including calculate the isotopic
pattern from a given chemical formula, determining the possible elemental
compositions for a given mass, and calculate the exact mass from a given
chemical formula.

The MolecularFormulaGenerator class was originally developed and evaluated 
as part of the MZmine~2 framework~\cite{Pluskal2012}. The following code
calculates all possible chemical formula for a given accurate mass, 
within allowed counts for each element:

@@

@@

  % \subsubsection*{SMARTS parser}

@@

@@

\subsubsection*{Molecular fingerprints}
Also the molecular fingerprints in CDK has seen development since the 1.2
release. Traditionally CDK fingerprints were represented using the
\textit{BitSet} class shipping with Java. The benefits of that class include
such things as already implemented methods for modifying bit sets using other
bit sets. However the class keeps a vector of bits in memory. The solution was
excellent for hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
\textit{i.e.}, 10 bits indexing space. However implementing a fingerprint
designed to avoid collisions with, \textit{e.g.} a 32 bit indexing space using
this approach would be highly memory-inefficient. To allow for multiple
fingerprint representation implementations a bit fingerprint interface was
introduced. Also, although fingerprints traditionally are bit vectors a count
fingerprint was also introduced making fingerprints based on integer vectors
supported in CDK as well. The fingerprints currently existing in CDK are listed
in Table~1\ref{tab:fingerprints}.

@@

An overview of key modules with description, important changes, and dependencies
on third-party libraries is given in Table~2\ref{tab:modules} and the dependencies
between the CDK modules are depicted in Figure~6\ref{fig:deps}.

@@

  \subsection*{Table 1 - The molecular fingerprints in CDK.}
  \label{tab:fingerprints}
  Listed are the currently available molecular fingerprint in CDK with
  information about whether they come as bit / count versions, what CDK version
  they appeared in, their default size and where applicable also relevant
  references.
  \vskip 1\baselineskip

@@

      \subsection*{Table 2 - A selection of key CDK modules with major changes.}\label{tab:modules}
  An overview of a selection of often used CDK modules with description,
  dependencies on third-party libraries, and the major changes since
  version~1.2. Dependencies between modules are depicted in Figure~4\ref{fig:deps}.
  \vskip 1\baselineskip

>>>>

% Load packages
\usepackage{url}  % Formatting web addresses  
\urlstyle{rm}
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amssymb}
%\usepackage{cite}
\usepackage{graphicx}
\usepackage{multirow}

@@

\author[
   addressref={um},                   % id's of addresses, e.g. {aff1,aff2}
   %corref={aff1},                       % id of corresponding address, if any
   %noteref={n1},                        % id's of article notes, if any
   email={egon.willighagen@maastrichtuniversity.nl}   % email address
]{\inits{ELW}\fnm{Egon L} \snm{Willighagen}}
\author[
   addressref={aff2},
   email={john@nextmovesoftware.com}
]{\inits{JWM}\fnm{John W} \snm{May}}
\author[
   addressref={uppsala},
   email={jonathan.alvarsson@farmbio.uu.se}
]{\inits{JA}\fnm{Jonathan} \snm{Alvarsson}}
\author[
   addressref={uppsala},
   email={berg.arvid@gmail.com}
]{\inits{AB}\fnm{Arvid} \snm{Berg}}
\author[
   addressref={jena},
   email={kai.duehrkop@uni-jena.de}
]{\inits{KD}\fnm{Kai} \snm{Dührkop}}
\author[
   addressref={idea},
   email={jeliazkova.nina@gmail.com}
]{\inits{NJ}\fnm{Nina} \snm{Jeliazkova}}
\author[
   addressref={wi_mit},
   email={pluskal@wi.mit.edu}
]{\inits{TP}\fnm{Tomáš}~\snm{Pluskal}}
\author[
   addressref={???},
   email={miguelrojasch@googlemail.com}
]{\inits{MRJ}\fnm{Miguel}~\snm{Rojas-Cherto}}
\author[
   addressref={uppsala},
   email={ola.spjuth@farmbio.uu.se}
]{\inits{OS}\fnm{Ola} \snm{Spjuth}}
\author[
   addressref={???},
   email={gilleain.torrance@gmail.com}
]{\inits{GT}\fnm{Gilleain} \snm{Torrance}}
\author[
   addressref={um},
   email={chris.evelo@maastrichtuniversity.nl}
]{\inits{CTE}\fnm{Chris T} \snm{Evelo}}
\author[
   addressref={nih},
   email={guhar@mail.nih.gov}
]{\inits{RG}\fnm{Rajarshi} \snm{Guha}}
\author[
   addressref={ebi},
   email={steinbeck@ebi.ac.uk}
]{\inits{CS}\fnm{Christoph}~\snm{Steinbeck}}

\address[id=um]{
  \orgname{Dept of Bioinformatics - BiGCaT, NUTRIM, Maastricht University}, % university, etc
  %\street{Waterloo Road},                     %
  \postcode{NL-6200 MD},
  \city{Maastricht},                              % city
  \cny{The Netherlands}                                    % country
}
\address[id=aff2]{
  \orgname{??},
  %\street{D\"{u}sternbrooker Weg 20},
  %\postcode{24105},
  %\city{Kiel},
  \cny{UK}
}
\address[id=uppsala]{
  \orgname{Department of Pharmaceutical Biosciences, Uppsala University},
  %\street{D\"{u}sternbrooker Weg 20},
  \postcode{751 24},
  \city{Uppsala},
  \cny{Sweden}
}
\address[id=jena]{
  \orgname{Chair for Bioinformatics, Friedrich Schiller University},
  %\street{},
  \postcode{07743},
  \city{Jena},
  \cny{Germany}
}
\address[id=idea]{
  \orgname{Ideaconsult Ltd},
  \street{A. Kanchev 4},
  \postcode{1000},
  \city{Sofia},
  \cny{Bulgaria}
}
\address[id=wi_mit]{
  \orgname{Whitehead Institute for Biomedical Research}, 
  \street{9 Cambridge Center},
  \postcode{MA 02142},
  \city{Cambridge},
  \cny{USA}
}
\address[id=nih]{
  \orgname{National Center for Advancing Translational Science},
  \street{9800 Medical Center Drive},
  \postcode{MD 20878},
  \city{Rockville},
  \cny{USA}
}
\address[id=ebi]{
  \orgname{Chemoinformatics and Metabolism team, European Bioinformatics Institute},
  %\street{A. Kanchev 4},
  %\postcode{1000},
  \city{Hinxton},
  \cny{UK}
}

%\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
%\end{artnotes}

@@

  A chemical formula is the basic/simple chemical representation of a compound.
It defines the number of isotopes or elements that constitute the composition
of a compound without describing how atoms are bonded. With the rise of
metabolomics it has become increasingly relevant to have full support for these
in cheminformatics libraries~\cite{RojasCherto2011}.

  The CDK interfaces can handle several concepts related to chemical formulas:
the formula itself, sets of formulas, chemical formula ranges, adducts, isotope
containers and patterns, and rules to filter formula sets. These new tools can
be used for a number of tasks, including calculating the isotopic pattern from
a given chemical formula, determining the possible elemental compositions for a
given mass (mass decomposition), and calculating the exact mass from a given
chemical formula.

  The CDK contains two algorithms for the decomposition of mass ranges into
possible elemental formulas.  For most inputs, a Round Robin algorithm,
originally developed for the SIRIUS metabolite identification
tool~\cite{Bocker2009}, is used. The algorithm discretizes the real-value mass
decomposition problem into an integer-value knapsack
problem~\cite{Martello1990}. It first computes a dynamic programming table and
then backtracks it to generate matching formulas~\cite{Duehrkop2013,
Boecker2008}. Data for the Round Robin algorithm is stored in an extended
residue table~\cite{Bocker2005}, resulting in a low memory footprint of several
kilobytes. For certain problem instances, such as very large mass values (above
$400\,000$ Da) or mass range span larger than $1$ Da, the Round Robin algorithm
is not applicable and CDK falls back to an optimized full enumeration search
method, originally developed as part of the MZmine~2 framework for mass
spectrometry data processing~\cite{Pluskal2012, Pluskal2010}.

  The following code calculates all possible chemical formulas for a given
accurate mass, within allowed counts for each element:

@@

   To evaluate the performance of the CDK molecular formula generator, we
compared its runtimes on a range of inputs and compared them to the runtimes of
the classic, full enumeration-based HR2 formula generator~\cite{Kind2007} and
those of a recently developed Parallel Formula Generator (PFG)~\cite{Zhang2016}
(Table~1\ref{tab:formula_generators}). Although for small mass values the PFG
algorithm achieved shorter runtimes, for larger masses the Round Robin
algorithm implemented in CDK clearly outperformed the two other methods.

@@

  \subsubsection*{Substructure searching}

  TODO: iterator and new SMARTS

@@

  \subsubsection*{MDL molfile improvements}

  TODO: S group, V3000 formats, ...

@@

\subsubsection*{Molecular fingerprints}
Also the molecular fingerprints in CDK has seen development since the 1.2
release. Traditionally CDK fingerprints were represented using the
\textit{BitSet} class shipping with Java. The benefits of that class include
such things as already implemented methods for modifying bit sets using other
bit sets. However the class keeps a vector of bits in memory. The solution was
excellent for hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
\textit{i.e.}, 10 bits indexing space. However implementing a fingerprint
designed to avoid collisions with, \textit{e.g.} a 32 bit indexing space using
this approach would be highly memory-inefficient. To allow for multiple
fingerprint representation implementations a bit fingerprint interface was
introduced. Also, although fingerprints traditionally are bit vectors a count
fingerprint was also introduced making fingerprints based on integer vectors
supported in CDK as well. The fingerprints currently existing in CDK are listed
in Table~2\ref{tab:fingerprints}.

@@

An overview of key modules with description, important changes, and dependencies
on third-party libraries is given in Table~3\ref{tab:modules} and the dependencies
between the CDK modules are depicted in Figure~6\ref{fig:deps}.

@@

  \subsection*{Table 1 - Evaluation of molecular formula generators.}
  \label{tab:formula_generators}
  The resulting formula counts and runtimes of the HR2, PFG, and CDK chemical
formula generators based on input mass and mass tolerance values. Formulas were
generated using chemical elements C,H,N,O,P,S without bounds (the allowed atom
count was set to 0--10000 for each element). All heuristic filtering rules were
disabled for the purpose of the evaluation. The slight differences in the
number of generated formulas were caused by different isotope masses embedded
in each software and/or by rounding errors during calculation. The runtimes are
average values from three independent runs performed on three different 16-core
Intel Xeon 2.2~GHz CPU workstations equipped with 128~GB RAM, running Ubuntu
Linux version 12.04.5 LTS and OpenJDK Java runtime version 1.7.0\_101.
  \vskip 1\baselineskip

    \begin{minipage}{1\textwidth}
    \centering
    \begin{tabular}{llllllll}
	\multirow{2}{*}{ Mass (Da) } & \multirow{2}{*}{ Tolerance ($\pm$ Da) } & \multicolumn{3}{c}{ \# of generated formulas } & \multicolumn{3}{c}{ Runtime (s) } \\
	& & HR2 & PFG & CDK & HR2 & PFG & CDK \\
	\multirow{2}{*}{ $100$ } & $0.001$ & $1$ & $1$ & $1$ & $0.10$ & \textbf{$0.06$} & $0.49$ \\
	& $0.01$ & $18$ & $18$ & $18$ & $0.03$ & \textbf{$0.01$} & $0.52$ \\
	\multirow{2}{*}{ $500$ } & $0.001$ & $430$ & $430$ & $430$ & $0.51$ & \textbf{$0.09$} & $0.67$ \\
	& $0.01$ & $4\,265$ & $4\,265$ & $4\,264$ & $0.55$ & \textbf{$0.39$} & $0.96$ \\
	\multirow{2}{*}{ $1\,000$ } & $0.001$ & $8\,572$ & $8\,572$ & $8\,568$ & $16.5$ & $1.5$ & \textbf{$1.2$} \\
	& $0.01$ & $85\,542$ & $85\,542$ & $85\,537$ & $16.8$ & $5.4$ & \textbf{$3.7$} \\
	\multirow{2}{*}{ $2\,000$ } & $0.001$ & $226\,346$ & $226\,346$ & $226\,323$ & $1156.5$ & $58.9$ & \textbf{$8.6$} \\
	& $0.01$ & $2\,263\,410$ & $2\,263\,410$ & $2\,263\,409$ & $932.5$ & $156.3$ & \textbf{$69.7$} \\
	\multirow{2}{*}{ $3\,000$ } & $0.001$ & $1\,647\,132$ & $1\,647\,132$ & $1\,647\,118$ & $9\,667.9$ & $603.5$ & \textbf{$59.3$} \\
	& $0.01$ & $16\,471\,338$ & $16\,471\,338$ & $16\,471\,373$ & $9\,694.5$ & $1\,303.6$ & \textbf{$505.7$} \\
    \end{tabular}
    \end{minipage}

  \subsection*{Table 2 - The molecular fingerprints in CDK.}
  \label{tab:fingerprints}
  Listed are the currently available molecular fingerprint in CDK with
  information about whether they come as bit / count versions, what CDK version
  they appeared in, their default size and where applicable also relevant
  references.
  \vskip 1\baselineskip

@@

      \subsection*{Table 3 - A selection of key CDK modules with major changes.}\label{tab:modules}
  An overview of a selection of often used CDK modules with description,
  dependencies on third-party libraries, and the major changes since
  version~1.2. Dependencies between modules are depicted in Figure~4\ref{fig:deps}.
  \vskip 1\baselineskip
